---
title: "NBA"
author: "Narcel Reedus"
date: "October 11, 2017"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load libraries into RStudio
```{r}
library(dplyr)
library(magrittr)
library(ggplot2)
```
## Read data
```{r}
NBA_train <- read.csv("C:/Users/narce/OneDrive/Documents/GitHub/NBA/NBA/NBA_train.csv")

NBA <-NBA_train
```
## Examine the structure of the data (835 observations and 20 variables)
```{r}
str(NBA)
```
## Variables definitions

## SeasonEnd: Year the season ended
## Team: Name of team
## Playoffs: Binary variable for playoff appearance
## W: Wins (regular season) 
## PTS: Points scored (regular season)
## oppPTS:Oopponent points scored (regular season)
## FG, FGA: Field goals (including three pointers)
## X2P, X2PA: 2-pointers
## X3P, X3PA: 3-pointers
## FT, FTA: Free throws
## ORB, DRB: Offensive and defensive rebounds
## AST: Assists
## STL: Steals
## BLK: Blocks
## TOV: Turnovers

## Determine how many regular season wins needed to make playoffs 
## by grouping the data by the number of wins and 
## creating three new features: number of playoff appearances, 
## and the percentage of playoff appearances fracPO divided by total number of wins

```{r}
tmp<- group_by(NBA, W) %>% summarise(nTot = n(), nPO = sum(Playoffs), fracPO = nPO/nTot)

View(tmp)
```
## Plot y=tmp$W, x=tmp$fracPO 

```{r}
plot(tmp$W, tmp$fracPO, pch = 21, col = "red2", bg = "orange", 
     xlab = "Wins", ylab = "Making the Playoffs", main = "Playoff Appearance / Regular Season Wins")
abline(h = 0.9, lty = 3, col = "red2")
abline(v = 35, lty = 3, col = "blue2")
abline(v = 45, lty = 3, col = "blue2")
```
## Results show that (45>) wins have a (90>) chance of making it to the playoffs
## Predicting wins by calculating the difference between points scored (PTS) and points allowed (oppPTS)
```{r}
NBA$PTSdiff <- NBA$PTS - NBA$oppPTS
```
## Check the linear relationship between PTSdiff and Wins

plot(NBA$PTSdiff, NBA$W, pch = 21, col = "red2", bg = "orange",
     xlab = "PTS Difference", ylab = "Wins", main = "Wins / Points Scored")


## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
